Against Spurious Understanding: Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-02-06 ECONDAT.tex
uh
improvement to tex setup

\begin{abstract}
Humans have a tendency to place a `human' like quality in the objects around them. Giving your car a name or talking to your pet, such that you believe they understand you in a human way. This behavior is also seeing traction in Machine Learning (ML), where the idea of sentience is being placed upon Large Language Models (LLMs). In this position paper, we discuss why humans invariably anthropomorphize onto various objects and how the search for Artificial General Intelligence (AGI) is pushing recent literature to over-emphasize sentience. To combat this, we provide a short study that investigates the latent space of models, demonstrating that the discovery of patterns in latent spaces should not be a surprising outcome. Finally, we discuss how the popularity in media and the race for AGI enforces over-interpretation of consciousness in the models we use.

\end{abstract}

\section{Introduction}
\label{submission}

% \textcolor{magenta}{TODOs for the team:
% \begin{itemize}
%     \item Please double check the citations. Note this style file does not retain capitalization in titles. A trick to retain capitalization is to put an extra set of brackets around the title. I did this for the Ladouceur et al. and Bubeck et al references already; please make this consistent throughout.
%     \item As another issue with the citations: note that arxiv preprints do not show very well. Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation. Please straighten this out for the other articles too
% \end{itemize}
% }

In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}. While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.

Generally, in the popularity of ChatGPT and the integration of generative AI in productivity tools (e.g.\ through Microsoft's Copilot integrations in GitHub and Office applications), one also can wonder whether the promise of AI is more in computationally achieving general intelligence, or rather in the engineering of general-purpose tools\footnote{A Swiss army knife is an effective general-purpose tool, without people wondering whether it is intelligent.}. Regardless, stakes and interests are high, e.g.\ with ChatGPT clearing nearly \$1 billion in months of its release\footnote{\url{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}}.

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.

Noticing these trends, we as the authors of this article are concerned. We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding. At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims. With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs). In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.

To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.

%Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI. As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer. With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.

%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}. This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{https://twitter.com/wesg52/status/1709551516577902782?s=20} (formally Twitter) platform. Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.

%This only gets amplified as the interest in Artificial General Intelligence has now merged with a push towards building general-purpose tools.

%\textcolor{magenta}{cynthia: I am now moving the texts of Andrew and Antony on AGI here; maybe some tightening can already be done in this. Unfortunately no time for thorough editing yet at this stage, but will come back to this later}

%In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'.

%Present work case studies an example, to which claims of observed AGI 'sparks' in AI research, and reviews work on two human phenomena that are likely to lead to false claims: anthropomorphism, or the attribution of human-like qualities to non-human objects, and confirmation bias, or the seeking-out and/or biased interpretation of evidence in support of one's beliefs. We describe how these phenomena may interact in the context of AI research and development, leading to especially poor inferences about the degree of general intelligence, or lack thereof, within AI systems. We further display results from experiments that at first appear to indicate nascent 'sparks' of AGI, but that stronger tests show are indications of spurious sparks.

%As previously mentioned, AGI is the current objective for which the global AI community is shooting for. As observed in 2023, the mainstream implementation of Large Language Models (LLMs) resulted in numerous societal impacts. This enabled those companies with the more mature models to reap considerable monetary returns, despite the numerous negative news feeds. True AGI will be another global milestone for AI and the entire scientific community, thus we already witness massive claims towards it.

% Recently, Google DeepMind made claims their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, which they state it does without the need for human demonstrations during training. Although this is a fantastic achievement by the team, where we observe the model outperforming previous state of the art (SOTA) models, the underlying model is far from AGI. As discussed in the paper \cite{trinh2024geometry}, AlphaGeometry utilizes a rule-bound deduction engine, allowing the use of an NLP model for further suggestions when it is unable to locate a logical answer. With this combination, it was able to score highly (\textit{25/30} questions solved) where the gold standard human equivalent was \textit{25.9/30}.

% However, as noted by Hector Zenil \cite{hector2024linkedin}, that work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry. They were able to achieve this success due to geometry being an axiomatically closed system. Therefore, despite DeepMinds success in creating a powerfully fast Geometry solving machine, it is still far from being an AGI.

%it is perhaps not surprising to see human-metaphors in how AI has been designed and described, even though these very designs prevent it from ever having the cognitive functions of a human e.g. \cite{salles2020anthropomorphism, van2023reclaiming}.

\section{Surprising Patterns in Latent Spaces?}\label{patterns-in-latent-spaces-and-how-to-find-them}

In Winter 2023, a research article went viral on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formerly Twitter) platform~\cite{gurnee2023languagev1}. In this paper, through linear probing experiments, the claim was made that LLMs learned literal maps of the world, and as such were more than `stochastic parrots'~\cite{bender2021dangers} that can only correlate and mimic existing patterns from data, but not understand truly understand it.

While the manuscript immediately received public criticism~\cite{marcus2023muddles}, and the subsequent, current manuscript version is more careful with regard to its claims~\cite{gurnee2023languagev2}, reactions on X seemed to largely exhibit excitement and surprise at the authors' findings. However, in this section, through various simple examples, we make the point that observing patterns in latent spaces should not be a surprising revelation. After starting with a playful example of how easy it is to `observe' a world model, we build up a larger example focusing on key economic indicators and central bank communications.

%One of the more contentious pieces of literature recently studied, is the paper by \cite{gurnee2023languagev2}. This paper, which was the inspiration for our work, received numerous points of critical feedback when they discussed their work on the X\footnote{\url{https://twitter.com/wesg52/status/1709551516577902782?s=20}} (formally Twitter) platform. Although not directly mentioning AGI, they do hint at the emergence of intelligence, specifically with the creation of the world model.

%They claim said models generate an internal representation of the world, with which spatiotemporal representations are created. Further to this we see claim to the generalization of tasks, in which the model can answer questions regarding spatial and temporal relationships, without the need for further training. Although these can be seen as building blocks for a more comprehensive model, it is also fair to assume that it is mere conjecture and more likely spurious correlations.

%In this section, we navigate a number of simple examples to support our claim, that observing patterns in latent spaces should not be a surprising revelation.
%To support our claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples. 
%We will draw on ideas from the fields of Economics and Finance, to illustrate further that this phenomenon is neither surprising nor unique to the field of computer science. We will start with very simple examples to demonstrate that even small and simple models can learn meaningful representations of the data. The final example in Section \ref{ex-llm} is more complex and closer in spirit to the experiments conducted by \citet{gurnee2023language}. As we progress, we will discuss both the benefits and potential pitfalls of finding patterns in latent spaces.

\begin{figure}

{\centering \includegraphics[width=0.5\textwidth]{results/figures/map.png}

}

\caption{\label{fig-map}Predicted coordinate values (in-sample) from a linear probe on final-layer activations of an untrained neural network.}

\end{figure}

\subsection{A Neural Networks Born with World Models?}\label{example-deep-learning}

\citet{gurnee2023languagev2} extract and visualize the alleged geographical world model by training linear regression probes on internal activations in LLMs (including Llama-2) for the names of places, to predict geographical coordinates associated with these places. Now, the Llama-2 model has ingested huge amounts of publicly available data from the internet, including Wikipedia dumps from the June-August 2022 period \citep{touvron2023llama}. It is therefore highly likely that the training data contains geographical coordinates, either directly or indirectly. At the very least, we should expect that the model has seen features during training that are highly correlated with geographical coordinates. The model itself is essentially a very large latent space to which all features are randomly projected in the very first instance before being passed through a series of layers which are gradually trained for downstream tasks.

In our first example in this section, we simulate this scenario, stopping short of training the model. In particular, we take the \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv} that was used in \citet{gurnee2023languagev2}, which maps locations/areas to their latitude and longitude. For each place, it also indicates the corresponding country. From this, we take the subset that contains countries that are currently part of the top 10 \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA world ranking}, and assign the current rank to each country (i.e., Argentina gets 1, France gets 2, \ldots{}).

Subsequently, to ensure that the training data only involves a noisy version of the coordinates, we transform the longitude and latitude data, respectively, as follows: \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\).

Next, we encode all features except the FIFA world rank indicator as continuous variables: \(X^{(n \times m)}\) where \(n\) is the number of samples and \(m\) is the number of resulting features. Additionally, we add a large number of random features to \(X\) to simulate the fact that not all features ingested by Llama-2 are necessarily correlated with geographical coordinates. Let \(d\) denote the final number of features, i.e.~\(d=m+k\) where \(k\) is the number of random features.

We then initialize a small neural network, which we consider as a \emph{projector}, that maps from \(X\) to a single hidden layer with \(h<d\) hidden units and sigmoid activation, and from there to a lower-dimensional output space. Without performing any training on the \emph{projector}, we simply compute a forward pass of \(X\) through the \emph{projector} and retrieve the activations \(\mathbf{Z}^{(n\times h)}\).

Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge regression: \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\) where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the longitude and latitude for each sample. A hold-out set is reserved for testing. Finally, we compute the predicted coordinates for each sample in the hold-out set as \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on a world map (see Figure~\ref{fig-map}).

While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand. Thus, with this simple example, we illustrate that meaningful target representations should be recoverable from a sufficiently large latent space, given the projection of a small number of highly correlated features.

Similarly, \citet{alain2018understanding} observe that even before training a convolutional neural network on MNIST data, the layer-wise activations can already be used to perform binary classification. In fact, it is well-known that random projections can be used for prediction tasks \cite{dasgupta2013experiments}.

\begin{figure*}[htp]

\centering{

\includegraphics[width=1.0\textwidth]{results/figures/pca.png}

}

\caption{\label{fig-pca}The first two principal components of US Treasury yields over time at daily frequency. Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}

\end{figure*}

\subsection{PCA as a Yield Curve Interpreter}\label{example-principal-component-analysis}
%\subsection{Example: Principal Component Analysis}\label{example-principal-component-analysis}

We now move to a concrete application domain: Economics. Here, the yield curve, plotting the yields of bonds against their maturities, is a popular tool for investors and economists to gauge the health of the economy.
%It plots the yields of bonds against their maturities.
The yield curve's slope is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy. %While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone.
To leverage this information in downstream modelling tasks, economists regularly use PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data (e.g.\ \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing}).

%Being primarily interested in understanding the nature of this low-dimensional projection, we limit our scope to the US Treasury yield curve. To this end, we
To understand the nature of this low-dimensional projection, we collect daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury. Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). Let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.

The top panel in Figure~\ref{fig-pca} shows the first two principal components of the yield curves of US Treasury bonds over time. Vertical stalks indicate key dates related to the Global Financial Crisis (GFC). During its onset, on 27 February 2007, financial markets were in turmoil following a warning from the Federal Reserve (Fed) that the US economy was at risk of a recession. The Fed later reacted to mounting economic pressures by gradually reducing short-term interest rates to unprecedented lows. Consequently, the average level of yields decreased and the curve steepened. Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively. In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.

%Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: 
Again, it should not be surprising that these latent embeddings are meaningful: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance after controlling for the effect of all previous components.

\begin{figure*}

\centering{

\includegraphics[width=1.0\textwidth]{results/figures/dl.png}[tb]

}

\caption{\label{fig-dl-results}The left chart shows the actual GDP growth
and fitted values from the autoencoder model. The right chart shows the
observed average level and spread of the yield curve (solid) along with
the predicted values (in-sample) from the linear probe based on the latent embeddings
(dashed).}

\end{figure*}%
%\subsection{Example: Autoencoder}\label{example-deep-learning}
\subsection{Autoencoders as Economic Growth Predictors}\label{example-deep-learning}

%So far we have considered simple linear transformations of data. One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning. Our next example will therefore involve a deep neural network.

Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks. In particular, we will now use a neural network architecture.
%, that allows us to recover a compact latent representation of the yield curve.

\subsubsection{Data}\label{data}

To estimate economic growth, we will rely on a quarterly
\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St.~Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we transform the data using log differences. Since
our yield curve data is daily, we aggregate it to the
quarterly frequency by taking averages of daily yields for each maturity. We also standardize yields since deep learning models tend to perform better with standardized data \cite{gal2019standardization}. Since COVID-19 was a substantial structural break in the time series, we also filtered out all observations after 2018.

\subsubsection{Model}\label{model}

%Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\). Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\). Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).

%To do this, we use a simple autoencoder architecture. With this
Using a simple autoencoder architecture, we let our model \(g_t\) denote growth and our conditional \(\mathbf{r}_t\) the matrix of aggregated Treasury yield rates at time \(t\). Finally, we let \(\theta\) denote our model parameters. Formally, we are interested in maximizing the likelihood \(p_{\theta}(g_t|\mathbf{r}_t)\).

%To do this, we use a simple autoencoder architecture: 
The encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer connecting the encoder to the decoder, is a fully connected layer with 6 neurons. The decoder consists of two fully connected layers, each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data. The output layer consists of a single neuron for our output variable, \(g_t\). We train the model over 1,000 epochs to minimize mean squared error loss using the Adam optimizer~\citep{kingma2017adam}.

The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
model. The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected. Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set. As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread. Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.

\subsubsection{Linear Probe}\label{linear-probe}

While the loss function applies most direct pressure on layers near the final output layer, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}. On a per-neuron basis, the pressure to distill useful representation is therefore likely maximized there. Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.  We compute and extract these activations \(A_t\) for all time periods \(t=1,...,T\). Next, we use a linear probe to regress the observed
yield curve factors on the latent embeddings. Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively. Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\), we then predict the yield curve factors% from the latent embeddings
: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).

The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve factors over time, while dashed lines show predicted values. We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread. %As in all of the other examples we have presented so, we do not find this result particularly surprising.

Did the neural network now learn an intrinsic understanding of the economic relationship between growth and the yield curve? To us, that would be too big of a statement. Still, the current form of information distillation can be useful, even beyond its intended use for monitoring models.

For example, an interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We might be interested in understanding to what degree economic growth in the past is associated with economic growth today. As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient. But this coefficient suffers from confounding bias since there are many other confounding variables at play, of which some may be readily observable and measurable, but others may not.

We e.g.\ already mentioned the relationship between interest rates and economic growth. To account for that, while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors. While this slightly changes the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant (column (2) in Table \ref{tab-reg}). This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth. Because we have included two additional regressors with little to no predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.

Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model. In particular, we pick the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings. The estimated coefficient on this latent factor is small in magnitude, but statistically significant. The overall model fit, as measured by the BIC has improved and the magnitude of the coefficient on lagged growth, has changed quite a bit. While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can be used in order to mitigate confounding bias.

\begin{table}
\caption{Regression output for various models.}\label{tab-reg}%
\input{results/regression.tex}
\end{table}

%\subsection{Example: Large Language Model}\label{ex-llm}
\subsection{LLMs for Economic Sentiment Prediction}\label{ex-llm}

So far we have considered simple linear transformations of data. One might argue that the previous examples do not really involve latent embeddings in the way that they are typically thought of in the context of deep learning. Our next example will therefore involve a deep-learning-based LLM.

Here, we closely follow the approach in \citet{gurnee2023languagev2} but apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset~\cite{shah2023trillion}. This dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences. The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'. To this end, they first manually annotate a sub-sample of the available data and then fine-tune various models for the classification task. Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.

To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'. In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy favors high interest rates to limit the supply of money and thereby control inflation. The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.

\subsubsection{Linear Probes}\label{linear-probes}
%Instead of computing a measure based on predicted labels,
We now use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators. Therefore, we further pre-process the data provided by \citet{shah2023trillion} and use their proposed model to compute activations of the hidden state, on the first entity token for each layer. We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.

For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity. To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels. To measure price inflatio,n we use percentage changes proxied by log differences.

To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations each time. In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer.
%We found that not doing so would generally lead to poor out-of-sample performance.
To account for stochasticity, we use an expanding window scheme with 5 folds for each indicator and layer. To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.

Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer. The values correspond to averages computed across cross-validation folds. Consistent with findings in related work \citep{alain2018understanding,gurnee2023languagev2}, we also observe that model performance tends to be higher for layers near the end of the transformer model. Curiously, for yields at longer maturities, we find that performance eventually deteriorates for the very final layers. We do not observe this for the training data, so we attribute this to overfitting.

% After all we are still regressing a single variable of interest on 128 regressors.

It should also be noted that performance improvements are generally of small magnitude. Still, the overall qualitative findings are in line with expectations. Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with the outcome of interest and achieve higher mean directional accuracy (MDA). Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.

To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models. For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10. These are not state-of-the-art forecasting models, but they serve as a reasonable baseline. We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures. After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform. Detailed results will be made available in a supplementary appendix.

\begin{figure*}

\centering{

\includegraphics[width=1.0\textwidth]{results/figures/rmse_pca_128.png}

}

\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators. The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}

\end{figure*}%

\begin{figure*}

\centering{

\includegraphics[width=1.0\textwidth]{results/figures/attack_all_measures.png}

}

\caption{\label{fig-attack}Probe predictions for sentences about inflation of prices (IP), deflation of prices (DP), inflation of birds (IB) and deflation of birds (DB). The vertical axis shows predicted inflation levels subtracted by the average predicted value of the probe for random noise.}

\end{figure*}%

%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
Even though \emph{FOMC-RoBERTa} (which is substantially smaller than the models tested in~\citet{gurnee2023languagev2} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions). So, have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
of the economy just by `reading' central bank communications? Should the FOMC use this LLM to direct their forward guidance?

% ICML sentence: Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?

%It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:

% \begin{quote}
% Have we uncovered further evidence that LLMs ``aren't mere stochastic
% parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
% of the economy just by `reading' central bank communications?
% \end{quote}

We are having a very hard time believing that the answer to either of these questions is `yes'. To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor. The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.

To test this, we select the best-performing probe trained on the final-layer activations for each indicator. We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about deflation in the number of birds. A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''. Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''. While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.

% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest. In other words, we conduct a small manual adversarial attack on the probe.

In light of the encouraging results for the probe in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category. If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing. More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.

To get to this last point, we also generate many probe predictions for samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote the linear probe that maps from the \(k\)-dimensional space spanned by \(k\) first principal components of the final-layer activations to the output variable of interest (CPI growth in this case). Then we sample \(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\) for \(i \in [1,1000]\) and compute the sample average. We repeat this process \(10000\) times and compute the median-of-means to get an estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\), that is the predicted value of the probe conditional on random noise.

Next, we propose the following hypothesis test as a minimum viable testing framework to assess if the probe results (may) provide evidence for an actual `understanding' of key economic relationships learned purely from text:

\begin{proposition}[Parrot
Test]\protect\hypertarget{prp-line}{}\label{prp-line}

~

>>>>

\begin{abstract}
\input{abstract_econdat.tex}
\end{abstract}

\input{body_econdat.tex}
